---
title: "Lotka-Volterra simulations"
author: "Adam Clark and Petr Keil"
date: "2023-12-06"
output: 
  html_document: 
    toc: yes
    highlight: tango
    theme: cerulean
    number_sections: yes
---

# Disclaimer

We've written this code on **Ubuntu Linux (20.04 LTS Focal Fossa)**. The code didn't work
on MS Windows, where the ODE solver did run into 32 vs 64 bit issues. 
We haven't tried on Mac OS. 

# Libraries and functions

```{r, warning = FALSE, message = FALSE}
rm(list=ls())
rm(list=ls())

library(deSolve) # for solving ODEs 
library(viridis)
library(tidyverse)
library(gridExtra)
library(randomForest)

system("R CMD SHLIB LVmod_metacom_log.c") # compile C code (needs Rtools)
source("LV_misc_functions.R")

set.seed(12345)
```

# Single run

## Set parameter values

```{r}
n0mu = 0.01 # mean starting abundance
n0sd = n0mu/2 # sd starting abundance
Kmu = 1 # mean K
Ksd = 0 # sd K
rmu = 1 # mean r
rsd = 0 # sd r
Amu = -0.8 # mean interaction strength
Asd = 0.2 # sd interaction strength
cmu = 0 # mean dispersal rate
csd = 0.2 # sd dispersal rate (among species)

minval = -999 # log(minval) is lower limit below which abundances are no longer tracked
# note - species only "die" in a cell if their abundance is driven to zero, but the -999 limit lets us avoid
# using lots of computational power to track infintesimally small abundances
M = 5 # number of sites
N0 = 12 # number of species (in global pool)
D = N0 # dimensionality of interaction matrix
steps = 20 # number of simulation steps

# make interaction matrix
Amat = matrix(nrow = N0, ncol = N0)
Amat[row(Amat)!=col(Amat)] = rnorm(N0^2-N0, Amu, Asd)
diag(Amat) = -1

# store prameters in a vector
Pars  <- c(N   = N0,                         # number of species
           M   = M,                          # number of patches
           minval = minval,                  # minimum population size to track
           K   = pmax(0, rnorm(N0, Kmu, Ksd)),   # vector of carrying capacities
           r  = rnorm(N0, rmu, rsd),         # vector of initial growth rates
           A  = c(Amat),                     # interaction coefficients
           d_c = 0.5,                          # proc. noise c
           d_z = 1,                          # proc. noise z
           # >2 means that more abundant species are hit more than less abund.
           d_n = 0,                          # proc. noise nugget
           d_m = 0,                          # proc. noise mean
           d_w = 1,                          # disturbance waiting time
           cv  = pmax(0, rnorm(N0, cmu, csd)),   # dispersal rate
           n0 = abs(rep(n0mu, N0)))          # initial abundance post-colonization

# process noise follows the formula:
#dsd = sqrt(dc*n^dz);
#n_new = n + (dsd+dn)*rnorm(1)+dm;

# make initial abundances
nini  <- abs(rnorm(N0*M, n0mu, n0sd))

# time to record abundances in simulation
times <- seq(0, steps, by = 1)

# log transform initial states - allows for more stable simulations
lnini = log(nini)
lnini[!is.finite(lnini)] = minval

# dummy variable for tracking disturbance times
lnini = c(lnini, 1, runif(1))

# dummy variable for tracking dispersal effects
lnini = c(lnini, rep(1, N0), runif(N0, 0.5, 1))
```

## Run the ODE solver

```{r}
dyn.load("LVmod_metacom_log.so") # load c code

#run C code
out_C <- ode(y = lnini, 
             times = times, 
             func = "derivs",
             parms = Pars, 
             dllname = "LVmod_metacom_log", 
             initfunc = "initmod",
             events = list(func = "event", root = TRUE), 
             rootfun = "myroot", 
             nout = N0, 
             nroot = N0+1)

dyn.unload("LVmod_metacom_log.so") # unload c code (helps with stablity)
```

## Explore the result

```{r}
# output: array with 3 dimensions (time, species, sites)
Mout = array(dim =c(nrow(out_C), N0, M))
Mout[] = exp(out_C[,1:(N0*M)+1])

# aggregate biomass across all local communities to the metacommunity scale
Mout.gamma <- apply(Mout, 1:2, sum)

# plot total biomass of each species at the metacommunity scale
matplot(Mout.gamma, type = "l")
abline(h=0)
```

# Main simulation runs

## Set simulation parameters

```{r}
params <- expand.grid(S = c(10, 100),
                      M = c(10, 100),
                      Amu = c(-0.9, -0.5, -0.1),
                      d_z = seq(1, 3, by = 0.2))
# repeat each combination of parameters rep.times
rep.times <- 10
params <- do.call("rbind", replicate(rep.times, params, simplify = FALSE))

# set times at which the community will be sampled
time1 = 10
time2 = 20

# set random seed
set.seed(12345)
```

## The main simulation loop

```{r}
#initialize results container
results <- list()

# the loop
for(i in 1:nrow(params))
{
  # print(paste("Simulation",i, "out of", nrow(params)))
  
  n0mu = 0.01 # mean starting abundance
  n0sd = n0mu/2 # sd starting abundance
  Kmu = 1 # mean K
  Ksd = 0 # sd K
  rmu = 1 # mean r
  rsd = 0 # sd r
  #Amu = -0.5 # mean interaction strength
  Amu = params$Amu[i]
  Asd = 0.2 # sd interaction strength
  cmu = 0 # mean dispersal rate
  csd = 0.2 # sd dispersal rate (among species)
  
  minval = -999 # log(minval) is lower limit below which abundances are no longer tracked
  # note - species only "die" in a cell if their abundance is driven to zero, 
  #        but the -999 limit lets us avoid
  # using lots of computational power to track infintesimally small abundances
  M = params$M[i]  # number of sites
  N0 = params$S[i] # number of species (in global pool)
  D = N0 # dimensionality of interaction matrix
  steps = 20 # number of simulation steps
  
  # make interaction matrix
  Amat = matrix(nrow = N0, ncol = N0)
  Amat[row(Amat)!=col(Amat)] = rnorm(N0^2-N0, Amu, Asd)
  diag(Amat) = -1
  
  # store prameters in a vector
  Pars  <- c(N   = N0,                         # number of species
             M   = M,                          # number of patches
             minval = minval,                  # minimum population size to track
             K   = pmax(0, rnorm(N0, Kmu, Ksd)),   # vector of carrying capacities
             r  = rnorm(N0, rmu, rsd),         # vector of initial growth rates
             A  = c(Amat),                     # interaction coefficients
             d_c = 0.5,                          # proc. noise c
             d_z = params$d_z[i],                          # proc. noise z
             d_n = 0,                          # proc. noise nugget
             d_m = 0,                          # proc. noise mean
             d_w = 1,                          # disturbance waiting time
             cv  = pmax(0, rnorm(N0, cmu, csd)),   # dispersal rate
             n0 = abs(rep(n0mu, N0)))          # initial abundance post-colonization
  
  # process noise follows the formula:
  #dsd = sqrt(dc*n^dz);
  #n_new = n + (dsd+dn)*rnorm(1)+dm;

  # make initial abundances
  nini  <- abs(rnorm(N0*M, n0mu, n0sd))
  nini  <- rlnorm(n = N0*M, meanlog = 1, sdlog = 1)
    
  # time to record abundances in simulation
  times <- seq(0, steps, by = 1)
  
  # log transform initial states - allows for more stable simulations
  lnini = log(nini)
  lnini[!is.finite(lnini)] = minval
  
  # dummy variable for tracking disturbance times
  lnini = c(lnini, 1, runif(1))
  
  # dummy variable for tracking dispersal effects
  lnini = c(lnini, rep(1, N0), runif(N0, 0.5, 1))
  
  dyn.load("LVmod_metacom_log.so") # load c code
  # run C code
  out_C <- ode(y = lnini, 
               times = times, 
               func = "derivs",
               parms = Pars, 
               dllname = "LVmod_metacom_log", 
               initfunc = "initmod",
               events = list(func = "event", root = TRUE), 
               rootfun = "myroot", 
               nout = N0, 
               nroot = N0+1)
  dyn.unload("LVmod_metacom_log.so") # unload c code (helps with stablity)
  
  # output: array with 3 dimensions (time, species, sites)
  Mout = array(dim =c(nrow(out_C), N0, M))
  Mout[] = exp(out_C[,1:(N0*M)+1])
  
  Mout.gamma <- apply(Mout, 1:2, sum)

  # ------------------------------------
  
  # extract Px and Ex using time1 and time2
  MAT <- Mout.gamma
  EXTract(MAT, time1, time2)
  PXTract(MAT, time1, time2)
  
  # number of extinctions at gamma scale
  E.gamma <- EXTract(Mout.gamma, 
                      time1, 
                      time2)
  # per-species prob of extinction at gamma scale
  P.gamma <- PXTract(Mout.gamma, 
                             time1, 
                             time2)
  
  # mean number of extinctions at alpha scale
  E.alpha <- mean(apply(X = Mout,
                            FUN = EXTract, 
                            MARGIN = 3, 
                            time1, 
                            time2))
  
  # per-species prob of extinction
  P.alpha <- mean(apply(X = Mout,
                         FUN = PXTract, 
                         MARGIN = 3, 
                         time1, 
                         time2))
  
  # slopes of PxAR and ExAR (note: slope = rise/run)
  P.slope <- (P.gamma-P.alpha) / (time2-time1)
  E.slope <- (E.gamma-E.alpha) / (time2-time1)
  
  results[[i]] <- c(params[i,],
                    P.alpha = P.alpha, 
                    P.gamma = P.gamma,
                    P.slope = P.slope,
                    E.slope = E.slope)
} # end of the main loop
```

## Process the simulations

```{r}
res <- as.data.frame(do.call("rbind", results))
res <- data.table::rbindlist(results)

# re-classify d_z to individual hypotheses
res$type[res$d_z < 2] <- "Negative DD"
res$type[res$d_z > 2] <- "Positive DD"
res$type[res$d_z == 2] <- "Constant death"

# save simulation results to a file
write.csv(res, file = "sim_ode_results.csv", row.names=FALSE)
```


## Visualize the simulation results

```{r}
# PROCESS AND VISUALIZE THE RESULTS OF THE SIMULATIONS

res <- read.csv("sim_ode_results.csv")

# plot the results
pxar.plot <- ggplot(data = res, aes(x = as.factor(d_z), y = P.slope)) +
  geom_boxplot(aes(fill = type)) + 
  #facet_grid(as.factor(S)~.) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  xlab(expression(d[z])) +
  ylab("PxAR slope") + 
  labs(title="(a)") +
  theme(legend.position = 'none')

exar.plot <- ggplot(data = res, aes(x = as.factor(d_z), y = E.slope)) +
  geom_boxplot(aes(fill = type)) +
  #facet_grid(as.factor(S)~.) +
  geom_hline(yintercept = 0, linetype = "dashed") +
    theme_classic() +
    xlab(expression(d[z])) +
    ylab("ExAR slope") + 
    labs(title="(b)") +
    theme(legend.position = c(0.9, 0.8))
  
grid.arrange(pxar.plot, exar.plot, nrow = 2)

pdf("Figure_PxAR_ExAR_from_LV_sims.pdf", width=8, height = 8)
  grid.arrange(pxar.plot, exar.plot, nrow = 2)
dev.off()

ggplot(data = res, aes(x = as.factor(d_z), y = E.slope)) +
  geom_boxplot(aes(fill = type)) + 
  #facet_grid(as.factor(S)~.) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  xlab(expression(d[z])) +
  ylab("PxAR slope") + 
  theme(legend.position = 'none') +
  facet_grid(.~M)
```

## Random forest analysis of the results

```{r}
# random forest explaining the PxAR slope
rf.P <- randomForest(P.slope ~ S + M + Amu + d_z,
                     data = na.omit(res)) # note the na.omit()

# variable importance
imp <- importance(rf.P)
imp <- data.frame(imp, predictor = rownames(imp))
imp <- imp[order(imp$IncNodePurity, decreasing=TRUE),]
imp$predictor <- with(imp, reorder(predictor, IncNodePurity, max))
imp$predictor <- factor(imp$predictor, levels = c("S","Amu","M","d_z"))


rf.imp.P <- ggplot(data = imp, aes(x = IncNodePurity, y = predictor)) +
            geom_col() + 
            xlab("Importance for PxAR slope") +
            ylab("Simulation parameter") + 
            labs(title="(a)") +
            theme_classic()
rf.imp.P

# -------------------------------------------------------

# random forest explaining the ExAR slope
rf.E <- randomForest(E.slope ~ S + M + Amu + d_z,
                     data = na.omit(res)) # note the na.omit()
# variable importance
imp <- importance(rf.E)
imp <- data.frame(imp, predictor = rownames(imp))
imp <- imp[order(imp$IncNodePurity, decreasing=TRUE),]
imp$predictor <- with(imp, reorder(predictor, IncNodePurity, max))
imp$predictor <- factor(imp$predictor, levels = c("S","Amu","M","d_z"))

rf.imp.E <- ggplot(data = imp, aes(x = IncNodePurity, y = predictor)) +
  geom_col() + 
  xlab("Importance for ExAR slope") +
  ylab("Simulation parameter") + 
  labs(title="(b)") +
  theme_classic()
rf.imp.E

grid.arrange(rf.imp.P, rf.imp.E, ncol=2, nrow = 1)

# ---------------------------------------

# export the figure to file
pdf("Figure_rf.pdf", width = 8, height = 4)
  grid.arrange(rf.imp.P, rf.imp.E, ncol=2, nrow = 1)
dev.off()

```


