---
title: "Code for Figure 1 illustrating the three main hypotheses"
author: "Petr Keil"
email: "keil@fzp.czu.cz"
date: "2023-11-28"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

This is a code that generates plots for Figure 1 in the main text. The figure
illustrates the three main hypotheses.

# Loading libraries, functions, and setting random seed

```{r, warning = FALSE, message = FALSE}
library(mobsim) # simulating point pattern communities

# graphics
library(tidyverse)
library(gridExtra) # multiple ggplots in a figure
library(latex2exp) # for math symbols in figures

# our own functions
source("1_point_pattern_functions.r")

# setting the random seed
set.seed(1234)
```

# Initial parameters 

```{r}
grains <- c(16,8,4,2)
dat <- SAD.prob.mass(1000, 100, cv_abund.vect = c(10))

N = 100 # number of individuals in the community
S = 10  # number of species
cv.abund = 1 # parameter of the SAD
sigma = 0.1 # spatial aggregation parameter
side.divisions = c(2,4,8,16) # four grid resolutions

# create the point pattern of the community
  com <- sim_thomas_community(n_sim  = N,
                              s_pool = S,
                              sad_coef = list(cv_abund = cv.abund),
                              mother_points = 1,
                              sigma  = sigma)

# extract N of individual species (this will be used in the Bartak function)
SAD <- data.frame(table(com$census$species))
names(SAD) <- c("species", "N.tot")
SAD$species <- as.character(SAD$species)
census <- left_join(com$census, SAD, by="species")
```

# Hypothesis 1 - Allee effect

## Bartak function

```{r}
# set parameters that correspond to the decreasing Bartak function (allee effect)
alpha = 0.9
beta = -1

bartak.allee.dat <- data.frame(N = seq(1,40), 
                              P.death = effect.f(alpha, beta, N = seq(1,40), plot=FALSE))

bartak.allee.plot <- ggplot(data = bartak.allee.dat, aes(x = N, y = P.death)) + 
        geom_point() +
        geom_line() +
        ylim(0,1) + 
        ylab(expression(P[death])) +
        theme_classic() + 
        xlab("N") 

bartak.allee.plot
```

## PxAR 

```{r}
# re-scale P.minus
P.minus.vect <- effect.f(alpha, beta, N = census$N.tot, plot = FALSE)

# sample individuals that will be killed
deaths <- rbinom(n = length(P.minus.vect), size = 1, prob = P.minus.vect)
dead.coords <- com$census[deaths == 1, ]

deltas <- com.extinction(com, 
                         side.divisions, 
                         alpha = alpha, 
                         beta = beta,
                         plot.res = FALSE)

PxAR.allee <- ggplot(data = deltas, aes(x = A, y = P.x)) + 
        geom_point() +
        scale_x_continuous(trans='log10') +
        geom_smooth(method = lm, se = FALSE, colour = "black", size = 0.5) +
        ylim(c(0,1)) + 
        theme_classic() +
        xlab("A") +
        ylab("Px")

PxAR.allee
```

## Point pattern grids

```{r, fig.width=10, fig.height = 4}
grids <- list()

for(i in 1:length(grains))
{
  grids[[i]] <-  ggplot(data = com$census, aes(x = x, y = y)) +
    geom_point(aes(colour = species)) + 
    geom_point(data = dead.coords, shape = 4, size = 2.5) +
    theme_minimal() + 
      coord_fixed(ratio = 1) +
      scale_x_continuous(breaks = seq(0, 1, by = 1/grains[i])) + 
      scale_y_continuous(breaks = seq(0, 1, by = 1/grains[i])) + 
      theme(axis.text.x=element_blank(),
            legend.position = "none",
            axis.text.y=element_blank(),
            panel.grid.major = element_line(color = "darkgrey"),
            panel.grid.minor = element_blank(),
            axis.title.x=element_blank(), axis.title.y=element_blank()) 
}

grid.arrange(grids[[1]], grids[[2]], grids[[3]], grids[[4]], nrow = 1)
```
  
Combined figure for Hypothesis 2
  
```{r}  
lay.mat <- matrix(c(1,1,2,2,
                    3,4,5,6), 
                  nrow = 2, ncol=4, byrow=TRUE)

 grid.arrange(bartak.allee.plot, PxAR.allee,
             grids[[1]], grids[[2]], grids[[3]], grids[[4]],
             layout_matrix = lay.mat)

pdf("Hypothesis_1.pdf", width = 7, height= 4.5)  
    grid.arrange(bartak.allee.plot, PxAR.allee,
               grids[[1]], grids[[2]], grids[[3]], grids[[4]],
               layout_matrix = lay.mat)
dev.off()
```

# Hypothesis 2 - Janzen-Connell effect

## Bartak function

```{r}
# set parameters that correspond to the decreasing Bartak function (allee effect)
alpha = 0.1
beta = 1

bartak.jc.dat <- data.frame(N = seq(1,40), 
                              P.death = effect.f(alpha, beta, N = seq(1,40), plot=FALSE))

bartak.jc.plot <- ggplot(data = bartak.jc.dat, aes(x = N, y = P.death)) + 
        geom_point() +
        geom_line() +
        ylim(0,1) + 
        ylab(expression(P[death])) +
        theme_classic() + 
        xlab("N") 

bartak.jc.plot
```

## PxAR 

```{r}
# re-scale P.minus
P.minus.vect <- effect.f(alpha, beta, N = census$N.tot, plot = FALSE)

# sample individuals that will be killed
deaths <- rbinom(n = length(P.minus.vect), size = 1, prob = P.minus.vect)
dead.coords <- com$census[deaths == 1, ]

deltas <- com.extinction(com, 
                         side.divisions, 
                         alpha = alpha, 
                         beta = beta,
                         plot.res = FALSE)

PxAR.jc <- ggplot(data = deltas, aes(x = A, y = P.x)) + 
        geom_point() +
        scale_x_continuous(trans='log10') +
        geom_smooth(method = lm, se = FALSE, colour = "black", size = 0.5) +
        ylim(c(0,1)) + 
        theme_classic() +
        xlab("A") +
        ylab("Px")

PxAR.jc
```

## Point pattern grids

```{r, fig.width=10, fig.height = 4}
grids <- list()

# loop through the four grids
for(i in 1:length(grains))
{
  grids[[i]] <-  ggplot(data = com$census, aes(x = x, y = y)) +
    geom_point(aes(colour = species)) + 
    geom_point(data = dead.coords, shape = 4, size = 2.5) +
    theme_minimal() + 
      coord_fixed(ratio = 1) +
      scale_x_continuous(breaks = seq(0, 1, by = 1/grains[i])) + 
      scale_y_continuous(breaks = seq(0, 1, by = 1/grains[i])) + 
      theme(axis.text.x=element_blank(),
            legend.position = "none",
            axis.text.y=element_blank(),
            panel.grid.major = element_line(color = "darkgrey"),
            panel.grid.minor = element_blank(),
            axis.title.x=element_blank(), axis.title.y=element_blank()) 
}

# plot the four grids
grid.arrange(grids[[1]], grids[[2]], grids[[3]], grids[[4]], nrow = 1)
```
  
Combined figure for Hypothesis 2
  
```{r}  
lay.mat <- matrix(c(1,1,2,2,
                    3,4,5,6), 
                  nrow = 2, ncol=4, byrow=TRUE)

grid.arrange(bartak.jc.plot, PxAR.jc,
             grids[[1]], grids[[2]], grids[[3]], grids[[4]],
             layout_matrix = lay.mat)

pdf("Hypothesis_2.pdf", width = 7, height= 4.5)  
  grid.arrange(bartak.jc.plot, PxAR.jc,
               grids[[1]], grids[[2]], grids[[3]], grids[[4]],
               layout_matrix = lay.mat)
dev.off()
```

# Hypothesis 3 - SAR, PxAR, and ExAR

```{r}
A = 1:100 # sequence of areas
# parameters of the SARs
c=3
z1 = 0.05
z2 = 0.1
z3 = 0.25
# slope of the PxAR
exz = -0.1
# vectors of species richness
S1 = c*A^z1
S2 = c*A^z2
S3 = c*A^z3

# CURVE 1 - shallow power law
Px = (c*A^exz)/max(c*A^exz)*0.5
Ex1 = S1*Px
Ex2 = S2*Px
Ex3 = S3*Px

# save everything to a data frame
dat <- data.frame(A = rep(A, times = 3), 
                  S = c(S1, S2, S3),
                  Px = rep(Px, times = 3),
                  Ex = c(Ex1, Ex2, Ex3),
                  scenario = rep(c("i","ii","iii"), each = length(A)))
```

Create the three panels

```{r}
SAR <- ggplot(data = dat, aes(x = A, y = S)) + 
  geom_line(aes(linetype = scenario)) + 
  labs(title = "SAR") +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  theme_classic()  +
  theme(legend.position="none")

SAR

PxAR <- ggplot(data = data.frame(A, Px), aes(x = A, y = Px)) + 
  geom_line() + 
  scale_x_continuous(trans='log10') +
  labs(title = "PxAR") +
  theme_classic()  +
  theme(legend.position="none")

PxAR

ExAR <- ggplot(data = dat, aes(x = A, y = Ex)) + 
  geom_line(aes(linetype = scenario)) + 
  scale_x_continuous(trans='log10') +
  labs(title = "ExAR") +
  ylab(expression(E[x]==S%*%P[x])) +
  theme_classic() +
  theme(legend.position="none")

ExAR

```  

Export the figures to a .pdf

```{r, fig.width = 10}
grid.arrange(SAR, PxAR, ExAR, nrow = 1)

pdf("Hypothesis_3.pdf", width=6, height=2.3)    
  grid.arrange(SAR, PxAR, ExAR, nrow = 1)
dev.off()
```
  

# Session info

```{r}
sessionInfo()
```





