---
title: "Main point pattern simulations"
author: "Petr Keil"
email: "keil@fzp.czu.cz"
date: "2023-11-28"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Settings and Libraries

```{r, warning = FALSE, message = FALSE}
library(mobsim) # simulating point pattern communities
library(manipulate) # sliders and interactive plotting
library(randomForest) # random forest

# graphics
library(tidyverse)
library(ggbreak)
library(gridExtra) # multiple ggplots in a figure
library(latex2exp) # for math symbols in figures

# parallel computation
library(foreach)
library(doParallel)

# our own functions
source("1_point_pattern_functions.r")

# setting the random seed
set.seed(12345)
```

# Histograms illustrating different SADs

```{r, fig.width=10}
dat <- SAD.prob.mass(1000, 100)
dat$cv_abund <- paste("CV[N]==", dat$cv_abund, sep="")

SADs <- ggplot(data = dat, aes(x=N, y = PM)) +
          geom_point() +
          geom_line() +
          theme_classic() + 
          ylim(0, 0.4) + ylab("Probablity mass") + 
          xlab("N") +
          labs(title="(c)") +
          facet_grid(.~cv_abund, labeller = label_parsed)
SADs
```

# Simulation of extinction in communities with different parameter values

```{r}
## set parameter values
params <- expand.grid(alpha = c(0.01, 0.1, 0.3, 0.5, 0.7, 0.9, 0.99),
                      beta = c(-4, -1, -0.4, 0, 0.4, 1, 4),
                      N.tot = c(100, 1000) , 
                      S.frac = c(0.05, 0.1, 0.2), 
                      sigma = c(0.01, 0.1, 1),
                      cv.abund = c(0.1, 1, 10))

params <- data.frame(params, S.tot = params$N.tot*params$S.frac)

# fix random seed
set.seed(12345)

# number of cores for the parallel simulation (each parameter combination
# will run N.cores times)
N.cores <- 10
params.list <- rep(list(params), times = N.cores)

# initialize cores
cl <- makeCluster(N.cores)
registerDoParallel(cl, cores = N.cores)

# run the simulation
results <- foreach(i = 1:N.cores, 
                   .combine = rbind,
                   .packages = c("mobsim", "tidyverse")) %dopar% 
{
  # the main.loop function that simulates the community,
  # executes sampling, and analyzes the scaling
  main.loop(params)
}

# release the cores
stopCluster(cl)

# classify results to the three types of N-P.death dependency
res <- data.frame(results, type = "")
res$type[res$beta < 0] <- "Negative DD"
res$type[res$beta > 0] <- "Positive DD"
res$type[res$beta == 0] <- "Constant death"
```

Export simulation results to a file:

```{r, eval=TRUE}
write.csv(res, file = "sim_results.csv", row.names=FALSE)
```

# Main plot of the beta-slope hypothesis

```{r}
res <- read.csv("sim_results.csv")
res$cv.abund <- paste("CV[N]==", res$cv.abund, sep="")

slope.vs.beta.E <- ggplot(data = res, aes(x = as.factor(beta), y = slope.E)) +
  #geom_vline(xintercept=as.factor(0), linetype = "dashed", colour = "grey") +
  geom_abline(intercept=0, slope = 0, linetype = "dashed", colour = "grey") +
  geom_boxplot(aes(fill = type), outlier.colour = "grey", outlier.shape = 1) +
  facet_grid(.~as.factor(cv.abund), labeller = label_parsed) + 
  theme_classic() +
  ylim(c(-1.2,1.2)) +
  labs(title="(b)") +
  xlab(expression(beta)) +
  ylab("ExAR slope") + 
  theme(legend.position = "none")
slope.vs.beta.E

slope.vs.beta.P <- ggplot(data = res, aes(x = as.factor(beta), y = slope.P)) +
  #geom_vline(xintercept=as.factor(0), linetype = "dashed", colour = "grey") +
  geom_abline(intercept=0, slope = 0, linetype = "dashed", colour = "grey") +
  geom_boxplot(aes(fill = type), outlier.colour = "grey", outlier.shape = 1) +
  facet_grid(.~as.factor(cv.abund), labeller = label_parsed) + 
  theme_classic() +
  xlab(expression(beta)) +
  ylab("PxAR slope") + 
  labs(title="(a)") +
  theme(legend.position = c(0.1, 0.8))
slope.vs.beta.P

slope.simple <- ggplot(data = res, aes(x = as.factor(beta), y = slope.P)) +
  #geom_vline(xintercept=as.factor(0), linetype = "dashed", colour = "grey") +
  geom_abline(intercept=0, slope = 0, linetype = "dashed", colour = "grey") +
  geom_boxplot(aes(fill = type), outlier.colour = "grey", outlier.shape = 1) +
  #facet_grid(.~as.factor(cv.abund)) + 
  theme_classic() +
  xlab(expression(beta)) +
  ylab("PxAR slope") + 
  theme(legend.position = c(0.8, 0.8))
slope.simple
```

Plot all of the results together

```{r, fig.width=10, fig.height=10}
grid.arrange( slope.vs.beta.P, slope.vs.beta.E, SADs, nrow = 3, heights=c(0.4, 0.7, 0.7))

## Export figure for the paper
pdf("Figure_beta_vs_scaling.pdf", width = 8, height=11)
  grid.arrange( slope.vs.beta.P, slope.vs.beta.E, SADs, nrow = 3, heights=c(0.7, 0.7, 0.4))
dev.off()
```

# Analyze the simulations with random forest

Which simulation parameters and community properties are most important
in driving the slope of the extinction scaling?

## Random forest - PxAR

```{r}
rf.P <- randomForest(slope.P ~ alpha + 
                               beta + 
                               N.tot + 
                               sigma + 
                               cv.abund +
                               S.tot, 
                   data = na.omit(res)) # note the na.omit()

# extract variable importance
imp <- importance(rf.P)
imp <- data.frame(imp, predictor = rownames(imp))
# order results by importance
imp <- imp[order(imp$IncNodePurity, decreasing=TRUE),]
# turn variable names into an ordered factor
imp$predictor <- with(imp, reorder(predictor, IncNodePurity, max))
imp$predictor <- factor(imp$predictor, levels = c("N.tot","S.tot", "sigma",
                                                  "cv.abund","alpha", "beta"))

# plot the importances
rf.imp.P <- ggplot(data = imp, aes(x = IncNodePurity, y = predictor))+
  geom_col() +   
  xlab("Importance for PxAR slope") +
  ylab("Simulation parameter") + 
  labs(title="(a)") +
  theme_classic()
rf.imp.P
```

# Random forest - ExAR

```{r}
rf.E <- randomForest(slope.E ~ alpha + 
                               beta + 
                               N.tot + 
                               sigma + 
                               cv.abund +
                               S.tot, 
                   data = na.omit(res)) # note the na.omit()

# extract variable importance
imp <- importance(rf.E)
imp <- data.frame(imp, predictor = rownames(imp))
# order results by importance
imp <- imp[order(imp$IncNodePurity, decreasing=TRUE),]
imp$predictor <- with(imp, reorder(predictor, IncNodePurity, max))
# turn variable names into an ordered factor
imp$predictor <- factor(imp$predictor, levels = c("N.tot","S.tot", "sigma",
                                                  "cv.abund","alpha", "beta"))
# plot the importances
rf.imp.E <- ggplot(data = imp, aes(x = IncNodePurity, y = predictor))+
  geom_col() +   
  xlab("Importance for ExAR slope") +
  ylab("Simulation parameter") + 
  labs(title="(b)") +
  theme_classic()
rf.imp.E
```

Export all importance figures to a file

```{r}
pdf("Figure_rf.pdf", width = 8, height = 4)
grid.arrange(rf.imp.P, rf.imp.E, nrow = 1, ncol = 2)
dev.off()
```

# Plots of effect of the variables found by the random forest analysis

```{r, fig.width = 10, fig.height = 5}
# effect of the mean probability of death on extinction scaling
ggplot(data = res, aes(x = mean.P.minus, y = slope.P)) +
  geom_abline(intercept=0, slope = 0) +
  geom_point(aes(colour = type), alpha = 0.5, shape = 1, size = 0.5) +
  theme_classic() +
  geom_smooth(se=FALSE, aes(colour = type), method=lm, formula = y ~ poly(x, 2)) +
  xlab(expression(paste("Average ", P[death]))) +
  ylab("PxAR slope") +
  geom_smooth(se=FALSE, aes(colour = type)) +
  facet_grid(.~cv.abund)

# effect of mean log abundance on extinction scaling
ggplot(data = res, aes(x = mean.log.abund, y = slope.P)) +
  geom_abline(intercept=0, slope = 0) +
  geom_point(aes(colour = type), shape = 1, alpha = 0.5, size = 0.5) +
  theme_classic() +
  xlab("Mean log Abundance") +
  ylab("PxAR slope") +
  geom_smooth(se=FALSE, aes(colour = type), method=lm, formula = y ~ poly(x, 2)) +
  facet_grid(.~cv.abund)

# effect of alpha on extinction scaling
ggplot(data = res, aes(x = as.factor(beta), y = slope.P)) +
  geom_abline(intercept=0, slope = 0) +
  geom_boxplot(aes(fill = type), outlier.colour = "grey") +
  facet_grid(as.factor(alpha)~as.factor(cv.abund)) + 
  theme_classic()
```

# Session info

```{r}
sessionInfo()
```
